Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-07 submission.tex
Tag set optimization

\paragraph{Tags \& Features} % <-- fixme: rename?
As we seek to quantify the effects of PoS tagging in a realistic setting, i.e.,
in application to raw text, we choose to evaluate the parser on automatically
assigned PoS tags. For training, however, we have two options: using either
gold standard or automatically assigned tags. In order to settle on a
configuration, we conducted experiments with training on both gold standard
tags and automatically assigned tags to see how the performance differs. The
results of these experiments, shown in Table \ref{tagconfigeval}, reveal that
the combination of training and testing on automatic tags is clearly superior
to training on gold standard tags and testing on automatic tags. Consequently,
the parser is both trained and tested on automatically assigned tags in the
remainder of our experiments.

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Training} & \textbf{Testing} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        Gold & Gold & 90.15\% & 92.51\% \\
        Gold & Auto & 85.68\% & 88.98\% \\
        Auto & Auto & 87.01\% & 90.19\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of parsing with Mate using the various tag configurations,
        either using gold standard tags or automatically assigned from TnT.
        \emph{Gold} denotes gold standard tags, \emph{Auto} denotes
        automatically assigned tags from TnT.}
    \label{tagconfigeval}
\end{table}

Though some parsers make use of morphological features, % fixme: status for mate?
we removed all morphological features beyond the PoS tags in order to simulate
a realistic setting.  Moreover, it is crucial that we remove these features
when working with automatically assigned tags, as such features would be
provided by the gold standard.

@@

In the following, we report on a set of experiments which evaluate increasingly
fine-grained PoS tag sets. The PoS tag sets are motivated by morphosyntactic
properties of Norwegian and are evaluated in terms of both tagger and parser
accuracy.

\section{Tag Set Optimization}
\label{sec:optimization}
By introducing more fine-grained linguistically motivated distinctions in a tag
set, we increase the linguistic information represented in the tags, which may
assist the parser in recognizing and generalizing syntactic patterns.
We would then also increase the complexity of the tagging task though,
which can be expected to lead to a drop in tagger accuracy.
However, the best tagging, evaluated in isolation, does not necessarily lead to the
best parse, hence it is interesting to investigate how tag set modifications
may affect this interplay.

Hence, we are in actuality altering the gold
standard tags by the addition of more relevant linguistic information,
producing new gold standard annotations of higher quality.

% fixme: tror nok dette bør flyttes og kortes ned. kan bakes inn med det som
% står under "Tag Set Mapping".
Our approach is somewhat semi-automatic in that we employ a "hybrid"
combination of manual annotation and automatic labeling in our experiments. Our
initial tags are gold standard, and using linguistic insight coupled with
computational considerations, we introduce new gold standard distinctions to
the tag set. The data sets with these tags are run through a pipeline with
automatically assigned tags in both training and testing, where the most
promising modifications are used "further". Hence, the tag set modifications
are not deterministic, nor do we try all possible combination; we only
experiment with linguistically motivated distinctions.

As we want to investigate how we can improve the linguistic quality of a tag
set, we will only consider tag set modifications we deem linguistically
sensible.

\subsection{Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its full set of morphological features, thereby mapping the original tag set to a
new maximally fine-grained tag set given the annotations available in the
treebank. This resulted in a total of 368 tags,
hereafter referred to as the \emph{full} tag set.
The two initial tag sets, i.e., the original tag set comprising 19 tags and the
full tag set comprising 368 tags, thus represent two extremes in terms of
granularity.
To establish some initial points of reference for how tag set granularity affects the
performance of tagging and parsing on NDT, we trained and tested a full pipeline with both of
these intital tags sets. The results are reported in Table \ref{inittagseteval}.
Unsurprisingly, we see that the tagger accuracy
plummets when we move from the original to the full tag set. While the MFT
tagging baseline for the original tag set is 94.14\%, it drops by almost 9
percentage points to 85.15\% for the full tag set. TnT reports an accuracy of
97.47\% on the original tag set, which is reduced to 93.48\% for the full tag
set. These results confirm our hypothesis that the very high level linguistic
information
in the full, fine-grained tag set comes at the expense of reduced tagger
performance.  In spite of this, however, we see that the additional information
in the full tag set still improves the parser performance. With the original tag set, Mate
yields 87.01\% LAS and 90.19\% UAS, which increases to 87.15\% and
90.39\%, respectively, using the full tag set. As we are looking for
linguistically informed distinctions that improve the syntactic parsing, these
results are promising and indicate that additional linguistic
information assists syntactic parsing, motivating the optimization of the
existing PoS tag set.

>>>>

\paragraph{Predicted vs gold tags}
As we seek to quantify the effects of PoS tagging in a realistic setting, i.e.,
in application to raw text, we choose to evaluate the parser on automatically
assigned PoS tags. For training, however, we have two options: using either
gold standard tags or automatically predicted tags. In order to settle on a
configuration, we conducted experiments with training on both gold tags 
and predicted tags to see how the performance differs. The
results of these experiments, shown in Table \ref{tagconfigeval}, reveal that
the combination of training and testing on predicted tags is clearly superior
to training on gold standard tags and testing on predicted tags. Consequently,
the parser is both trained and tested on automatically predicted tags in the
remainder of our experiments.\footnote{Though some parsers also make use of
  morphological features, we removed all morphological features beyond the PoS
  tags in order to simulate a realistic setting.}

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Training} & \textbf{Testing} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        Gold & Gold & 90.15\% & 92.51\% \\
        Gold & Auto & 85.68\% & 88.98\% \\
        Auto & Auto & 87.01\% & 90.19\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of parsing with Mate using various configurations of PoS
      tag sources in training and testing: \emph{Gold} denotes gold standard
      tags and \emph{Auto} denotes tags automatically predicted by TnT.}
    \label{tagconfigeval}
\end{table}

% Moreover, it is crucial that we remove these features
% when working with automatically assigned tags, as such features would be
% provided by the gold standard.

@@

% In the following, we report on a set of experiments which evaluate increasingly
% fine-grained PoS tag sets. The PoS tag sets are motivated by morphosyntactic
% properties of Norwegian and are evaluated in terms of both tagger and parser
% accuracy.

\section{Tag Set Optimization}
\label{sec:optimization}

The modified tag sets used in the experiments reported in this section are
defined as combinations of PoS tags and morphological features already
available in the gold annotations of the treebank. In effect, we redefine the
tags comprising the gold standard as provided by NDT. The best performing
configuration can be seens as tag set specifically optimized for dependency
parsing. Note that the tag selection process itself can be seen as
semi-automatic; while we for each introduced distinction empirically evaluate
its impact on parsing performance -- as to see whether it is worth including in
the final configuration -- the process is manually guided by linguistic
considerations regarding which combinations to evaluate in the first place.
Although our expressed goal is to identify a tag set optimized for the
downstream task of dependency parsing, we will only consider tag set
modifications we deem linguistically sensible.

By introducing more fine-grained linguistically motivated distinctions in the tag
set, we increase the linguistic information represented in the tags, which may
assist the parser in recognizing and generalizing syntactic patterns.
However, this would also increase the complexity of the tagging task, 
which can be expected to lead to a drop in tagger accuracy.
However, the best tagging, evaluated in isolation, does not necessarily lead to the
best parse, hence it is interesting to investigate how tag set modifications
affect this interplay.

% Hence, we are in actuality altering the gold
% standard tags by the addition of more relevant linguistic information,
% producing new gold standard annotations of higher quality.

% fixme: tror nok dette bør flyttes og kortes ned. kan bakes inn med det som
% står under "Tag Set Mapping".
% Our approach is somewhat semi-automatic in that we employ a "hybrid"
% combination of manual annotation and automatic labeling in our experiments. Our
% initial tags are gold standard, and using linguistic insight coupled with
% computational considerations, we introduce new gold standard distinctions to
% the tag set. The data sets with these tags are run through a pipeline with
% automatically assigned tags in both training and testing, where the most
% promising modifications are used "further". Hence, the tag set modifications
% are not deterministic, nor do we try all possible combination; we only
% experiment with linguistically motivated distinctions.

\subsection{Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its full set of morphological features, thereby mapping the original tag set to a
new maximally fine-grained tag set, given the annotations available in the
treebank. This resulted in a total of 368 tags,
hereafter referred to as the \emph{full} tag set.
The two initial tag sets, i.e., the original tag set comprising 19 tags and the
full tag set comprising 368 tags, thus represent two extremes in terms of
granularity.
To establish some initial points of reference for how tag set granularity affects the
performance of tagging and parsing on NDT, we trained and tested a full pipeline with both of
these intital tags sets. The results are reported in Table \ref{inittagseteval}.
Unsurprisingly, we see that the tagger accuracy
plummets when we move from the original to the full tag set. While the MFT
tagging baseline for the original tag set is 94.14\%, it drops by almost 9
percentage points to 85.15\% for the full tag set. TnT reports an accuracy of
97.47\% on the original tag set, which is reduced to 93.48\% for the full tag
set. These results confirm our hypothesis that the very high level linguistic
information
in the full, fine-grained tag set comes at the expense of reduced tagger
performance.  In spite of this, however, we see that the additional information
in the full tag set still improves the parser performance. With the original tag set, Mate
yields 87.01\% LAS and 90.19\% UAS, which increases to 87.15\% and
90.39\%, respectively, using the full tag set. As we are looking for
linguistically informed distinctions that improve the syntactic parsing, these
results are promising and indicate that additional linguistic
information assists syntactic parsing, motivating the optimization of the
existing PoS tag set.
